---
title: "Sparrow M3C3_Nimbl"
author: "Martinius T. Singdahlsen"
date: "2025-04-15"
output: html_document
---

## Setup of packages, the R-markdown document and data frames

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                     cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")

library(nimble)
library(tidyr) #Has replace_na() func
library(dplyr)
library(jagsUI)
library(tidybayes)
library(brms)
library(latex2exp)
library(ggplot2)
library(GGally)
library(boot) #for logit function
```

### Data frame creation

## The capture histories

```{r}
#Df of all observations
s.df <- read.table('C:\\Users\\Martinius T.S\\Documents\\MSMNFMA\\Master related\\Saved data\\df sparrow\\all_inds_juv_ad.txt',sep=",", header = 1)

head(s.df)

#Df of only island of interest
assign("s27.df",subset(s.df,s.df$Island== 27))
s27.df$date <- as.Date(s27.df$date, format="%Y-%m-%d") #Set it to date and not character

#Df of only island of interest with own column for just date in year
assign("s27.sep.df",s27.df)
s27.sep.df$time <- as.Date(format(s27.sep.df$date, "%m-%d"),format="%m-%d") # This is all year 2025?!
s27.sep.df$year <- as.numeric(format(s27.sep.df$date, "%Y"))
#s27.sep.df <- subset(s27.sep.df,select= -c(date))

#Everything above --> with no DNA observations
assign("s27.filled.df",subset(s27.sep.df,s27.sep.df$filled== "no"))
#s27.filled.df <- replace_na(s27.filled.df,list(sex="u")) We want NA as sex
```

## Morphology


```{r}


morph.df <- read.table('C:\\Users\\Martinius T.S\\Documents\\MSMNFMA\\Master related\\Saved data\\df sparrow\\morphology.csv',sep=",", header = 1)
head(morph.df)

#assign("morph27.df",subset(morph.df,morph.df$flok== 27)) #This restricts to island 27
morph.df <- morph.df %>%  # We set filed observation of sex to a number
  mutate(fieldsex = recode(fieldsex,"f"=1,"u"=0,"m"=2,"pm"=0,"pf"=0))
morph.df <- morph.df %>% # Set NA to zero
  mutate(fieldsex = ifelse(is.na(fieldsex), 0, fieldsex))

morph.df <- morph.df %>% # We set dna observation of sex to a number
  mutate(scriptsex = recode(scriptsex,"f"=1,"u"=0,"m"=2,"pm"=0,"pf"=0))
morph.df <- morph.df %>% # Set NA to zero
  mutate(scriptsex = ifelse(is.na(scriptsex), 0, scriptsex))

time.cov.df <- morph.df
time.cov.df$flok <- NULL
time.cov.df$flokb <- NULL
time.cov.df$fieldsex <- NULL
time.cov.df$scriptsex <- NULL
time.cov.df$scriptage <- NULL
time.cov.df$stage <- NULL
time.cov.df$init <- NULL
time.cov.df$hs_id <- NULL
time.cov.df$dato <- as.Date(time.cov.df$dato, format="%d/%m/%Y")
time.cov.df$year <- as.numeric(format(time.cov.df$dato, "%Y"))
time.cov.df$census <- time.cov.df$year - 1992 #So that first census is 1

#time.cov.df$tars_h[!is.na(time.cov.df$tars_h)]
norm_tand <- function(df,var){
  non.na.df <- df[[var]][!is.na(df[[var]])]
  mean_df <- mean(non.na.df)
  sd_df <- sqrt(var(non.na.df))
  return((df[[var]] - mean_df)/sd_df)
}

no.na.tars_h <- time.cov.df$tars_h[!is.na(time.cov.df$tars_h)]
time.cov.df$tars_h <- (time.cov.df$tars_h - mean(no.na.tars_h))/sqrt(var(no.na.tars_h))

no.na.nebb_h <- time.cov.df$nebb_h[!is.na(time.cov.df$nebb_h)]
time.cov.df$nebb_h <- (time.cov.df$nebb_h - mean(no.na.nebb_h))/sqrt(var(no.na.nebb_h))

no.na.nebb_l <- time.cov.df$nebb_l[!is.na(time.cov.df$nebb_l)]
time.cov.df$nebb_l <- (time.cov.df$nebb_l - mean(no.na.nebb_l))/sqrt(var(no.na.nebb_l))

no.na.ving_h <- time.cov.df$ving_h[!is.na(time.cov.df$ving_h)]
time.cov.df$ving_h <- (time.cov.df$ving_h - mean(no.na.ving_h))/sqrt(var(no.na.ving_h))

no.na.vekt <- time.cov.df$vekt[!is.na(time.cov.df$vekt)]
time.cov.df$vekt <- (time.cov.df$vekt - mean(no.na.vekt))/sqrt(var(no.na.vekt))

head(time.cov.df)

hist(time.cov.df$tars_h, xlim = c(-5,5),breaks = 30)
hist(time.cov.df$nebb_h, xlim = c(-5,5),breaks = 30)
hist(time.cov.df$nebb_l, xlim = c(-5,5),breaks = 30)
hist(time.cov.df$ving_h, xlim = c(-5,5),breaks = 30)
hist(time.cov.df$vekt, xlim = c(-5,5),breaks = 30)
```

Dato: Thats data
flok:
flokb: 
ringnr: ID
fieldsex: The sex it is visually seen as in the field
scriptsex: The sex it is DNA tested as
scriptage: 
tars_h: Bird statistic
nebb_h: Bird statistic
nebb_l: Bird statistic
ving_h: Bird statistic
vekt: Bird statistic
stage: Only "capt"
init: Initials of who measured
hs_id: 0, 1 or NA

We can remove, scriptage.

scriptage i think is information that is used in the field and has to do with the date the bird was first captured or its age. That information is in the other data frame. 

## Data reformating


```{r}
#Note that df used has no na's !! AND WE ARE USING DF.FILLED !!
#We make a df of the sparrow's, on long format. Then we can transform it to a standard upper diagonal design using previous code i develouped for the Stellar sea lions
s27.filled.df <- s27.filled.df %>%  # We set filed observation of sex to a number
  mutate(sex = recode(sex,"f"=1,"u"=5,"m"=0,"pm"=5,"pf"=5))
#s27.filled.df <- s27.filled.df %>% # Set NA to zero
#  mutate(sex = ifelse(is.na(sex), 0, sex))
s27.filled.df <- s27.filled.df %>%
  mutate(sex = na_if(sex, 5))

unique.id <- sort(unique(s27.filled.df$ID)) #List of all ID's
n.indi <- length(unique.id)

ch.df <- data.frame()
x.data.array <- array(NA,dim=c(5,31,n.indi)) #The covariates

census.count <- length(unique(s27.filled.df$year)) #How many censuses we have
year.min <- min(unique(s27.filled.df$year)) #Finding first year, 1993 #Dont think it is in use
i <- 1
for (var.id in unique.id){ #Here we make the data matrices for the model
  ### The CH df ###
  all.obs.df <- s27.filled.df[s27.filled.df$ID == var.id,]#All observations of the an individual
  years.obs <- unique(all.obs.df$year) #The years we observed the individual
  index.obs <- years.obs - (1993) + (1) #Here 1993 is 1 born
  first.co <- min(index.obs) #Finding first census observed
  last.co <- min(max(index.obs+4),30) #Last cohort we simulate ### Here the integer is how much further we simulate the individual from last observation 
  ch <- rep(0,30)
  ch[index.obs] <- 1 #Make ch of individual
  birth.year <- min(all.obs.df$Born_Year) #Birth year from df (it is a given covariate)
  birth.cohort.ind <- birth.year - (1993) + (1)
  min.observed.age <- min(all.obs.df$Age) #This is the workaround implementation
  sex_num <- max(all.obs.df$sex)
  
  ### Making X ###
  all.obs.x.df <- time.cov.df[time.cov.df$ringnr == var.id,]
  all.obs.x.df <- all.obs.x.df[order(all.obs.x.df$dato,decreasing = FALSE),] #The newest within year covariates are used, from oldest to newest
  cr <- unique(all.obs.x.df$census) #The unique years of observation
  census.range <- intersect(cr,index.obs) #To make certain it is observed when we have covariates. This is as the bird might emigrate and then having x.covariates is wrong.
  for(unique.census in census.range){#We look at each year with data for individual
    sightings.census <- all.obs.x.df[all.obs.x.df$census == unique.census,]
    for(index.covariate in 3:7){ #Here the 3:7 is index of c(tars_h,nebb_h,nebb_l,ving_h,vekt)
      covariate <- sightings.census[,index.covariate] #We assign the newest covariate observed within year
      index.first.non.NA <- which(!is.na(covariate))[1]
      eraliest.non.NA <- covariate[index.first.non.NA]
      x.data.array[index.covariate-2,unique.census,i] <- eraliest.non.NA #Assign covariate
    }
  }
  
  x.data.array[1,31,i] <- var.id #For data manipulation only
  x.data.array[2,31,i] <- birth.cohort.ind #For data manipulation only
  
  ### Setting everything together ###
  ch.df <- rbind(ch.df,c(ch,first.co,birth.cohort.ind,min.observed.age,sex_num,var.id,last.co))
  i <- i + 1
  
  
  if(i %% 500 == 0){
    cat(" ",round(i/n.indi,2)," ")
  }
}
print("Completed")


names.vec <- c()   #Here we give names to the columns
for (i in c(1:census.count)){names.vec <-c(names.vec,paste0("C",i))}
names(ch.df) <- c(names.vec,"first_ch","birth_cohort.ind","min_observed_age","sex","id","Last_obs5")

#saveRDS(x.data.array,file="x_simp_array.Rda")
#saveRDS(ch.df,file="ch_simp_array.Rda")

#Checking
#5422 observed only at c30
#3 Male
#7 Female and first observed at age 1

x.data.array[,,16]
time.cov.df[time.cov.df$ringnr=="8118442",]
identical(x.data.array[1,31,],ch.df[,35]) #True is returned
```

```{r}
### As there is no if statements in jags or Nimble. We further want to manipulate the data such that we don't have individuals that do not contribute to the likelihood, i.e they are only observed the last census. We also want individuals that are observed the next last census to be a special case as they only have one hidden survival node and not the rest
old.size <- dim(ch.df)[1]
for(i in c(1:old.size)){#Remove individuals first observed at ch 30
  j <- old.size - i + 1 
  if(ch.df[j,31]=="30"){
    ch.df <- ch.df[-c(j),]
    x.data.array <- x.data.array[,,-c(j)]
  }
}

#Checking
new.size <- dim(ch.df)[1]
dim.x <- dim(x.data.array)
dim.ch <- dim(ch.df)
print(dim.x)
print(dim.ch)
print(old.size - new.size) #This number checks out with what was found in M2
#5685 Is the amount of unique individuals found in 

cat("number of individuals,",dim.ch[1],". Amount of unique ID's,", length(unique(ch.df[,35])),". Amoun of non uniqe ID's,",length(ch.df[,35]))
identical(x.data.array[1,31,],ch.df[,35])
```


```{r}
###Now we simply sort by first observed
ordered.ch <- ch.df[order(as.double(as.double(ch.df[,32])),ch.df[,35]),] #Order first by cohort.ind then ID
ordered.x <- x.data.array[,,order(as.double(x.data.array[2,31,]),x.data.array[1,31,])] #Same here ^

#Checking
df1 <- as.data.frame(ordered.ch)
ordered.x[,,242]
ordered.x[,,243]
df1[242,]
df1[243,]
identical(ordered.x[1,31,], ordered.ch[,35]) #They are ordered the same way and some visual checks has been done
```


```{r}
#The data is correctly made now. We only need to make 1 alive and 2 dead because of the cat density in Nimble, and overwrite [1,1,] and [2,1,] in ordered.x

#ordered.x[1,1,] <- rep(NA,dim(ordered.x)[3]) #Overwrite temporary ID
#ordered.x[2,1,] <- rep(NA,dim(ordered.x)[3]) #Overwrite temporary first ch

x.final <- array(as.double(ordered.x[,-c(31),]), dim = dim(ordered.x[,-c(31),]))
ch.temp <- ordered.ch
ch.temp[,c(1:34,36)] <- lapply(ch.temp[,c(1:34,36)], as.double)
ch.temp[,1:30] <- ch.temp[,1:30] * (-1) + 2  #Transform to follow dcat(), i.e 1 alive 2 dead
ch.final <- ch.temp  #Not transform the extra statistics 

x.final[,,1117]
ch.final[1117,]
x.final[,,2024]

```

# Finding missing data index

```{r}
#Handling temporal data that is censored
#We only need to identify the data that are not observed. However Both observed and not observed data is found here.

missing.data.finder <- function(x.data,vec.cohort,vec.sim.end){ #Obs, code will crash if first individual has no observations of covariates or only has observation of covariates
  t.max <- dim(x.data)[2]
  
  temp <- x.data[,,1]
  if(vec.cohort[1]!=0){temp[,1:cohort] <- 1} #Special case of cohort zero. We need the covariates in the first column to predict survival
  if(vec.sim.end[1]!=30){temp[,vec.sim.end[1]:t.max] <- 1}
  na_indices <- cbind(which(is.na(temp), arr.ind = TRUE),1) #Here na_indices is defined first
  
  temp2 <- x.data[,,1] #This is quite redundant code. Temp1 and 2 is defined the same and if statment is quite similar
  if(vec.cohort[1]!=0){temp2[,1:cohort] <- NA}
  if(vec.sim.end[1]!=30){temp2[,vec.sim.end[1]:t.max] <- NA}
  real_indices <- cbind(which(!is.na(temp2), arr.ind = TRUE),1) #Same as in forloop
  
  n.itter <- length(vec.cohort)
  for(i in 2:n.itter){
    cohort <- vec.cohort[i]
    sim.end <- vec.sim.end[i]
    temp <- x.data[,,i]
    if(cohort!=0){temp[,1:cohort] <- 1} #No observation for zero year of life
    if(sim.end!=30){temp[,sim.end:t.max] <- 1} #Overwrite 1 where we don't continue the HMM because of not observing individual for long
    if(dim(which(is.na(temp2), arr.ind = TRUE))[1]!=0){ #If there are no NA's it crashes
      which.na <- which(is.na(temp), arr.ind = TRUE)
      col.i.na <- rep(i,dim(which.na)[1])
      na_indices <- rbind(na_indices,cbind(which.na,col.i.na))
    }
    temp2 <- x.data[,,i]
    if(cohort!=0){temp2[,1:cohort] <- NA} #No observation for zero year of life
    if(sim.end!=30){temp2[,sim.end:t.max] <- NA}
    if(dim(which(!is.na(temp2), arr.ind = TRUE))[1]!=0){ #If there are no Reals it crashes
      which.real <- which(!is.na(temp2), arr.ind = TRUE)
      col.i.real <- rep(i,dim(which.real)[1])
      real_indices <- rbind(real_indices,cbind(which.real,col.i.real))
    }
    if(i %% 500 == 0){
      cat(" ",round(i/n.itter,2)," ")
    }
  }
  return(list(na = na_indices,observe = real_indices))
}

index <- missing.data.finder(x.final,ch.final[,32],ch.final[,36])
index.non.last <- missing.data.finder(x.final,ch.final[,32],rep(30,5573))

#testing
#x.final[,,1] #Birth cohort 0 individual
#ch.final[1,]
#index$index.na

#index$index.real[2625:2650,] #It skipps individual that has no observations
#x.final[,,1005]

dim(index$na)
dim(index$observe)

dim(index.non.last$na)
dim(index.non.last$observe)

head(index$observe)
head(index$na)

#All the rows in a, not found in b
a <- index.non.last$observe
b <- index$observe

a[!duplicated(rbind(b,a))[(nrow(b)+1):(nrow(a)+nrow(b))], ] #Its empty which is good. This was solved by introducing the union poerator in x.data cration

```


```{r}
#Handling non temporal data that is NA, i.e sex
sex.na <- which(is.na(ch.final[,34]), arr.ind = TRUE)
sex.observe <- which(!is.na(ch.final[,34]), arr.ind = TRUE)
```

```{r}
# Here is how many hidden nodes we used to simulate compared to how many we now do simulate with the approximation of assuming that an individual is dead after not observing it for 4 years (or how many census we eventually choose).
sum(30-(ch.final[,32]-1))
sum(ch.final[,36]-(ch.final[,32]-1))

cat("How many precent of z nodes remain afther approximation of death:",sum(ch.final[,36]-(ch.final[,32]-1))/sum(30-(ch.final[,32]-1)))

#The way x.data is used now, the data from the census it is born is not used. I thus make a statistic of how much data we are not using

cohort.reduced <- ch.final[,32][-c(1:76)] #Remove zeroth cohort
x.reduced <- x.final[,,-c(1:76)]
na.count <- 0

for(i in 1:length(cohort.reduced)){
  cohort.ind <- cohort.reduced[i]
  na.count <- na.count + sum(is.na(x.reduced[,cohort.ind,i]))
  
}
cat("
")
cat("How many precent all covariates are censored of census that is the cohort of the corresponding individual:",na.count/(length(cohort.reduced)*5))

#Example of this
#x.final[,,5354]
#head(cov.na.500[1700:3000,])
```


## Fitting individual model


```{r}

N <- dim(ch.final)[1]
time <- 30

### 1000 different for censoring indexing ###
#Here we assume that the arrays of indexings are sorted by individual
index.na <- which(index$na[,3]>(N-501))[1] #Finding index of individ N-500
index.observe <- which(index$observe[,3]>(N-501))[1] #Finding index of individ N-500

cov.na.500 <- index$na[index.na:dim(index$na)[1],] #Inex collection of last 500 NA
cov.observe.500 <- index$observe[index.observe:dim(index$observe)[1],] #Inex collection of last 500 observed

#This one with sex I remember is kinda confusing. However I think it does not need to be to well understood as it will be removed for the final model
N.sex.na.500 <- length(which(is.na(ch.final[(N-500):N,34]), arr.ind = TRUE))
N.sex.observe.500 <- length(which(!is.na(ch.final[(N-500):N,34]), arr.ind = TRUE))

sex.na.500 <- sex.na[(length(sex.na)-N.sex.na.500):length(sex.na)]
sex.observe.500 <- sex.observe[(length(sex.observe)-N.sex.observe.500):length(sex.observe)]

### Latent variable model index setup ###
#beta.order.cov.na.500 <- cov.na.500[order(cov.na.500[,1]),] Just incase i need it
individual.na.500 <- cov.na.500[,-c(1)] #Removing covariate, now this is purely a matrix of individual since morphological data is assumed to be non temporal
individual.na.500 <- unique(individual.na.500) 


### Z ###
Z <- array(NA,dim=c(N,30))
for (i in c((N-500):N)){ ### 1000 Different
  Z[i,c((ch.final[i,32]+1):30)] <- 1
}

### W ###
W <- array(NA,dim=c(5,30,N))
for (i in c((N-500):N)){ ### 1000 Different
  W[,c((ch.final[i,32]+1):30),i] <- 0
}

### Prob_init ###
prob_init <- rep(0,time)
prob_init[c(1:26)] <- NA ### 1000 Different
prob_init_one <- prob_init + 1
prob_init_neg_one <- prob_init - 1

### Tau_init ###
#Precision matrix of corr=1/3 with unit variance
Tau_init <- diag(7)
Tau_init_low_corr <- diag(7)
Tau_init_neg_low_corr <- diag(7)

### L_init ###
L_init <- matrix(0,nrow=5,ncol=2)
L_init_pluss <- L_init + 1
L_init_minus <- L_init - 1

### f_init ###

f_init <- array(NA, dim = dim(x.final)[c(2,3)]) #This is starting values for one latent variable
for(i in 1:dim(individual.na.500)[1]){
  f_init[individual.na.500[i,1],individual.na.500[i,2]] <- 0
}
f_init_final <- array(c(f_init, f_init), dim = c(dim(f_init),2)) #We take two coppies and glue them together in to get three dimensional starting values for latent variables

f_init_final_pluss <- f_init_final + 1
f_init_final_minus <- f_init_final - 1

##Deal with initial values
inits.1 <-  list(prob_norm = prob_init, #N*1
       a = 0,
       b = 1,
       
       sex.prob = 1/2,
       
       Tau_beta = Tau_init, # p*p
       mu_beta = rep(0,7), # p*1
       beta = rep(0,7), # p*1
       
       L = L_init,
       L_tau = 2,
       f = f_init_final_minus,
       individual_tau_lmv = rep(2,5),
       
       z = Z) #N*time

inits.2 <-  list(prob_norm = prob_init_one,
       a = -1,
       b = 1/2,
       
       sex.prob = 3/4,
       
       Tau_beta = Tau_init_low_corr,
       mu_beta = rep(-1,7),
       beta = rep(1,7),
       
       L = L_init_pluss,
       L_tau = 1/2,
       f = f_init_final_pluss,
       individual_tau_lmv = rep(1,5),
       
       z = Z)

inits.3 <-  list(prob_norm = prob_init_neg_one,
       a = 1,
       b = 2,
       
       sex.prob = 1/4,
       
       Tau_beta = Tau_init_neg_low_corr,
       mu_beta = rep(1,7),
       beta = rep(-1,7),
       
       L = L_init_pluss,
       L_tau = 1,
       f = f_init_final,
       individual_tau_lmv = rep(1/2,5),
       
       z = Z)


initsLIST = list(inits.1,
                inits.2,
                inits.3)



##Data
Modeldata <- list(y=ch.final[,-c(31:35)],
                  x=x.final, 
                  sex = ch.final[,34])

modelConst <- list(N=N,time=30,
                   cohort.ind=ch.final[,32],
                   cohort.end=ch.final[,36], #zeroth.cohort=76,
                   
                   beta_n = 7,
                   mu_beta_mu = rep(0,7),
                   mu_beta_var = diag(7),
                   Identity=diag(7),
                   
                   X.na=cov.na.500, X.na.length=dim(cov.na.500)[1],
                   X.individual=individual.na.500, 
                   X.individual.length=dim(individual.na.500)[1], #This is all "individuals" (t and n) that have censored covariates vec.with.beta.index
                   
                   sex.na=sex.na.500,
                   na.sex.length=length(sex.na.500))
                   
#                   factor_mu=rep(0,2), #Why did this not work?
#                   factor_tau=diag(2))
### 1000 different                   
#                   na.index=dim(index$na)[1],
#                   X.na=index$na,
                   
#                   na.sex.index=length(sex.na),
#                   sex.na=sex.na,

#                   zeroth.cohort = 76)

          

#parameters to save
params <- c("rho","varians",                #Density approximation of precision
            "mu_beta",# "Tau_beta",               #Precision of cohort.ind survival rate
            "beta",                    #The cohort.ind survival
            "prob_norm","a","b",       #Observe prob
            "sex.prob",
            "L","L_tau","individual_tau_lmv")      #Prior for sex hyperparameter

```


```{r}
beta.index=vec.with.beta.index
X.na=beta.order.cov.na.500
mat <- matrix(0,nrow=1,ncol=3)
for(i in 1:5){
  for(indecies in (beta.index[i]+1):beta.index[i+1]){ 
    mat <- rbind(mat,c(X.na[indecies,1],X.na[indecies,2],X.na[indecies,3]))
  } 
}

identical(mat[-c(1),],X.na)
```




```{r}

N <- dim(ch.final)[1]
time <- 30

model.hidden <- nimbleCode({
  a ~ dnorm(0,1) #Hyperprior for prob detection
  b ~ dexp(1) #Hyperprior for prob detection
  sex.prob ~ dbeta(1,1) #Hyper Prior for Age
  
  mu_beta[1:beta_n] ~ dmnorm(mu_beta_mu[1:beta_n], mu_beta_var[1:beta_n,1:beta_n]) #Hyper prior for mean of beta
  Tau_beta[1:beta_n,1:beta_n] ~ dwish(Identity[1:beta_n,1:beta_n],beta_n+1) #Hyper prior for mean of beta
  beta[1:beta_n] ~ dmnorm(mu_beta[1:beta_n],Tau_beta[1:beta_n,1:beta_n]) #Beta
  

  L_tau ~ dexp(1) 
  for(index in 1:X.na.length){ # X.na[index,1] is p, X.na[index,2] is t, X.na[index,2] is i
    x[X.na[index,1],X.na[index,2],X.na[index,3]] ~ dnorm(lvm_mu[X.na[index,1],X.na[index,2],X.na[index,3]],individual_tau_lmv[X.na[index,1]])
    # x[p,t,i] ~ dnorm(lvm_mu[p,t,i],lvm_sigma[p]) one can think of t,i making up n
    lvm_mu[X.na[index,1],X.na[index,2],X.na[index,3]] <- L[X.na[index,1],1]*f[X.na[index,2],X.na[index,3],1]+L[X.na[index,1],2]*f[X.na[index,2],X.na[index,3],2]
    #lvm_mu[p,t,i] <- L[p,1]*f[t,i]+L[p,2]*f[t,i]
  }
  for(index in 1:X.individual.length){
    f[X.individual[index,1],X.individual[index,2],1] ~ dnorm(0,1) #Prior for factors
    f[X.individual[index,1],X.individual[index,2],2] ~ dnorm(0,1)
  }
  for(index in 1:5){
    L[index,1] ~ dnorm(0,L_tau) #Prior for factor loadings
    L[index,2] ~ dnorm(0,L_tau)
    individual_tau_lmv[index] ~ dexp(1)
  }


  
  for(index in 1:na.sex.length){ #Handeling of censoring of sex
    sex[sex.na[index]] ~ dbinom(size=1, prob=sex.prob)
  }
  
  for(t in 27:time){ #Heirarchial system for prob
    prob_norm[t] ~ dnorm(a,b)
    logit(prob[t]) <- prob_norm[t]
    
    Lambda[1,1,t] <- prob[t]      # Pr(alive t -> observe alive t)
    Lambda[1,2,t] <- 1 - prob[t]  # Pr(alive t -> observe dead t)
    Lambda[2,1,t] <- 0             # Pr(dead t -> observe alive t)
    Lambda[2,2,t] <- 1             # Pr(dead t -> observe dead t)
    
  }
  '
  for(i in 1:zeroth.cohort){#Zeroth cohort special case
    z[i,cohort.ind[i]] <- 1 #Initialization with prob of detecting
    #No observations were done in year 1992, thus no observation
    
    
    
    for(t in (cohort.ind[i] + 1):cohort.end[i]){
      logit(phi[i,t]) <- beta[1] + beta[2]*sex[i] + beta[3]*x[1,t,i] + beta[4]*x[2,t,i] + beta[5]*x[3,t,i] + beta[6]*x[4,t,i] + beta[7]*x[5,t,i] 
      
      Gamma[1,1,i,t] <- phi[i,t]      # Pr(alive t-1 -> alive t)
      Gamma[1,2,i,t] <- 1 - phi[i,t]  # Pr(alive t-1 -> dead t)
      Gamma[2,1,i,t] <- 0             # Pr(dead t-1 -> alive t)
      Gamma[2,2,i,t] <- 1             # Pr(dead t-1 -> dead t)
    
      z[i,t] ~ dcat(Gamma[z[i,t-1],1:2,i,t])  #Here z[i,t-1] is previous hidden state
      y[i,t] ~ dcat(Lambda[z[i,t],1:2,t])
    }
  }
  '
  
  for(i in (N-500):N){ #Likelihood and hierarchical system of the probability     t0 <- cohort.ind[i] + 1

    z[i,cohort.ind[i]] <- 1 #Initialization with prob of detecting, Could this be cat(c(1,0))? or ber(1)?
    y[i,cohort.ind[i]] ~ dcat(Lambda[z[i,cohort.ind[i]],1:2,cohort.ind[i]])
    
    for(t in (cohort.ind[i] + 1):cohort.end[i]){
      logit(phi[i,t]) <- beta[1] + beta[2]*sex[i] + beta[3]*x[1,t,i] + beta[4]*x[2,t,i] + beta[5]*x[3,t,i] + beta[6]*x[4,t,i] + beta[7]*x[5,t,i] 
      
      Gamma[1,1,i,t] <- phi[i,t]      # Pr(alive t-1 -> alive t)
      Gamma[1,2,i,t] <- 1 - phi[i,t]  # Pr(alive t-1 -> dead t)
      Gamma[2,1,i,t] <- 0             # Pr(dead t-1 -> alive t)
      Gamma[2,2,i,t] <- 1             # Pr(dead t-1 -> dead t)
    
      z[i,t] ~ dcat(Gamma[z[i,t-1],1:2,i,t])  #Here z[i,t-1] is previous hidden state
      y[i,t] ~ dcat(Lambda[z[i,t],1:2,t])
    }
  }

  #Estimation of Covariance
  Sigma[1:7,1:7] <- inverse(Tau_beta[1:7,1:7])
  for (k in 1:7){
    for (l in c(1:7)){
      rho[k,l] <- Sigma[k,l]/sqrt(Sigma[k,k]*Sigma[l,l])
    }
    varians[k] <- sqrt(Sigma[k,k])
  }
  
})


start_time <- proc.time()
mcmc.output.probNphi <- nimbleMCMC(code = model.hidden,
                          data = Modeldata,
                          constants = modelConst,
                          inits = initsLIST,
                          monitors = params,
                          thin = 5,
                          niter = 10000,
                          nburnin = 5000,
                          nchains = 3)
end_time <- proc.time()
print(end_time-start_time)
```

```{r}
#write.csv(mc.df, "C:\\Users\\Martinius T.S\\Documents\\MSMNFMA\\Master related\\Saved data\\.csv")

mc1.df <- as.data.frame(mcmc.output.probNphi$chain1)
mc2.df <- as.data.frame(mcmc.output.probNphi$chain2)
mc3.df <- as.data.frame(mcmc.output.probNphi$chain3)
names(mc1.df) <- paste(names(mc1.df),1,sep="")
names(mc2.df) <- paste(names(mc2.df),2,sep="")
names(mc3.df) <- paste(names(mc3.df),3,sep="")
mc.df <- cbind(mc1.df,mc2.df,mc3.df)
head(mc.df)

mc.df$`prob_norm[27]1` <- inv.logit(mc.df$`prob_norm[27]1`)
mc.df$`prob_norm[27]2` <- inv.logit(mc.df$`prob_norm[27]2`)
mc.df$`prob_norm[27]3` <- inv.logit(mc.df$`prob_norm[27]3`)

mc.df$`prob_norm[28]1` <- inv.logit(mc.df$`prob_norm[28]1`)
mc.df$`prob_norm[28]2` <- inv.logit(mc.df$`prob_norm[28]2`)
mc.df$`prob_norm[28]3` <- inv.logit(mc.df$`prob_norm[28]3`)

mc.df$`prob_norm[29]1` <- inv.logit(mc.df$`prob_norm[29]1`)
mc.df$`prob_norm[29]2` <- inv.logit(mc.df$`prob_norm[29]2`)
mc.df$`prob_norm[29]3` <- inv.logit(mc.df$`prob_norm[29]3`)

mc.df$`prob_norm[30]1` <- inv.logit(mc.df$`prob_norm[30]1`)
mc.df$`prob_norm[30]2` <- inv.logit(mc.df$`prob_norm[30]2`)
mc.df$`prob_norm[30]3` <- inv.logit(mc.df$`prob_norm[30]3`)

mc.df$index <- 1:dim(mc.df)[1] 
head(mc.df)

df_long <- pivot_longer(mc.df,cols = c('L[1, 1]1',"L[1, 1]2","L[1, 1]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("L[1, 2]1","L[1, 2]2","L[1, 2]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("beta[1]1","beta[1]2","beta[1]3"), names_to = "series", values_to = "value")


df_long <- pivot_longer(mc.df,cols = c("beta[1]1","beta[1]2","beta[1]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("beta[2]1","beta[2]2","beta[2]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("beta[3]1","beta[3]2","beta[3]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("beta[4]1","beta[4]2","beta[4]3"), names_to = "series", values_to = "value")

df_long <- pivot_longer(mc.df,cols = c("beta[7]1","beta[7]2","beta[7]3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("a1","a2","a3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("b1","b2","b3"), names_to = "series", values_to = "value")
df_long <- pivot_longer(mc.df,cols = c("prob_norm[29]1","prob_norm[29]2","prob_norm[29]3"), names_to = "series", values_to = "value")


ggplot(df_long, aes(x = 1:3000, y = value, color = series)) +
  geom_line() +
  theme_minimal()


ggplot(df_long, aes(x = value, color = series)) +
  geom_density(size = 1.2) +
  theme_minimal()



```


